### 2016-12-02 chapter1.tex
Corrections made to Experiment/Theory Sections

\section{Theory and Observations}
Much work has been done over the past several decades to try to understand the complicated picture of plasma structuring in the polar cap, including theoretical analysis of plasma dispersion relations, simulations of plasma dynamics, and observations of plasma structuring using a variety of instruments.

\subsection{Theory and Simulations}

Polar patches are large regions of enhanced plasma density in the polar F region \citep[][e.g.]{Weber1984,Buchau1983}.  They are know to emerge on the dayside and then propagate with the background plasma convection across the polar cap to the nightside, where they either disintegrate or recombine with enhanced-density plasma in the auroral oval \textbf{CITE SOMETHING}.

There have been at least 6 mechanisms proposed for how polar patches form on the dayside.
\begin{enumerate}
	\item Discrete changes in solar wind parameters, usch as IMF, density, speed, and pressure \citep{Sojka1994}
	\item Average flow patterns varying in time \citep{Anderson1988}
	\item Transient magnetopause reconnection \citep{Lockwook1992b}
	\item Plasma production by cusp particle precipitation \citep{Rodger1994,Millward1999}
	\item Plasma flow jet channels that cut continuous streaches of plasma into segments 	\citep{Valladares1998}
	\item Alfven wave coupling \citep{Prikryl1999}
\end{enumerate}
Although particle precipitation has been shown to create enhanced plasma density in the cusp region \citep{Roger1994}, it is unreasonable for this mechanism to account for the largest density enhancements observed in polar patches.  These high densities must originate from the resivor of solar illuminated plasma on the dayside.  Additionally, a mechanism is still required to separate patches from the cusp region.  High speed flow channels can ``cut'' segments off of a tongue of high-density plasma that has been pulled into the polar cap by convection \citep{Valladares1994,Valladares1998}.  Models have also shown changing large scale convection patterns quickly in time does produce density structures similar to patches \citep{Anderson1988}.  However, the mechanism that seems to be dominant for the majority of polar patches observed is transient magnetic reconnection \citep{Carlson2012}.

In order to discuss patch formation by transient magnetic reconnection, it is important to first understand the steady-state ionospheric convection patterns in the polar cap and how they relate to the magnetosphere.  Figure \ref{fig:magnetosphere} is a schematic of how the IMF, magnetosphere, and ionosphere are all interconnected.  In general, magnetic field lines in the magnetosphere can either be closed, meaning they connect only the north and south pole of the earth, or open, when one of the ends of the field lines originates from the sun such that the field line is connected directly to the IMF.  The polar cap boundary is generally defined as the border between open and closed field lines, shown by the dark line in figure \ref{fig:magnetosphere}.  By Faraday's Law, the total EMF around the polar cap boundary is equal to the rate of change of magnetic flux, equation \ref{eqn:pc_imf}.  In the polar cap, the magnetic field strength is relatively constant at around \(5\times 10^-5\) T, so any changes in magnetic flux must be due to changes in the polar cap area, \(A_{pc}\) due to the polar cap boundary expanding or contracting \citep{Lockwood1992a}.
\begin{equation}
	\label{eqn:pc_imf}
	-E = \frac{d\Phi_B}{dt} = B\frac{dA_{pc}}{dt}
\end{equation}
Magnetic reconnection often occurs on both the dayside and nightside of the magnetosphere, changing the amount of open magnetic flux, so the polar cap boundary can be highly dynamic.

@@

Magnetic reconnection between the IMF and the Earth's magnetosphere occurs along the X-line on the dayside magnetopause (blue line in figure \ref{fig:magnetosphere}).  The X-line in the magnetopuase can be mapped along the magnetic field lines to the ionosphere, where it is referred to as the merging gap (pink line in figure \ref{fig:magnetosphere}).  During reconnection events, the merging gap moves equatorward as the open magnetic flux into the polar cap increases \citep{Lockwood1992a}.  In addition, the potential along the X-line also gets mapped the the merging gap, assuming there is minimal drop in potential along the magnetic field lines \citep{Lockwood1992b}.  The combination of these two effects results in flux being transferred across the boundary at a rate equal to the applied voltage  along the X-line and hence plasma flow across the merging gap \citep{Lockwood1992b}, as seen in figure \ref{fig:patch_formation}.  As the reconnection event continues, the flow across the merging gap creates a convection pattern, which starts to pull high density plasma from the dayside into the polar cap.  If this continues consistently, eventually it can create a "tongue" of high-density plasma stretching between the dayside merging gap and the center of the polar cap, often called the tongue of ionization.  However, when the reconnection burst ends, the merging gap begins to relax back towards the polar cap and brings the dense plasma with it, figure \ref{fig:patch_formation}.  Before the flow stops completely, the patch "pinches off" of the ionization tounge and becomes an isolated island of high density plasma in the polar cap.  The merging gap returns to its original position and the entire process can repeat with the next reconnection event.  A series of reconnection events will create a string of isolated polar patches propigating through the center of the polar cap \citep{Lockwook1992b}.

@@

Plasma structuring occurs in the ionosphere when conditions are such that wave growth occurs.  Because different background conditions change factors such as the density of both neutrals and ions, the magnetic field strength, and so forth, the collison frequency, gyroradii, and ion inertial effects change at different altitudes.  This causes a variety of different instability mechanisms to be operational depending on the background conditions.  The Farley-Buneman instability (FBI), or the modified two-stream instability tends to be a major factor in the E region where ion inertial effects are high \citep{Farley1963,Buneman1963}.  Also operational in the E region but more dominant at higher altitudes is the gradient-drift instability (GDI) \citep{Simon1963,Hoh1963,Linson1970}.  When GDI is stable, wave growth can still occur sometimes along the magnetic field, which is known as the current-conductive instability (CCI).  Additionally, there are shear-driven processes, known as the Kelvin-Helmholtz instability (KHI) \textbf{CITE???}.

@@

Some of the first 1D investigations of the linear GDI growth rate found that it can be described by a simple expression involving the plasma drift velocity, \(V_E\), and the gradient scale length, \(L\) \citep{Simon1963,Hoh1963,Linson1970}.
\begin{equation}
	\label{eqn:gdi_old}
	\gamma = \frac{V_E}{L}
\end{equation}
This expression is only valid if the density gradient is in the same direction as the plasma drift velocity and wavevector direction is not considered.  The results of \citet{Keskinen1982,Keskinen1983} do consider the wavevector, \(\vec{k}\), to some extent, but still assume a general relationship between density gradients and plasma drift \citep{Tsunoda1988}.
\begin{equation}
	\label{eqn:gdi_80s}
	\gamma = \frac{k_y}{k}\frac{\vec{k}\cdot\vec{V}_E}{kL}
\end{equation}
The directional dependance of GDI has been shown to be significant, such that the growth rate can actually change signs (determining whether wave growth or damping occurs) with different wavevector directions, so it is important to consider this factor \citep{Makarevich2014c}.  In addition, these expressions are only applicable in a particular altitudinal regieme, usually the F region.

Recently, a new expression has been introduced that considers an arbitrary altitude, wavevector, and density gradient within the ionosphere, which allows FBI, GDI, an CCI to be considered with a single dispersion relation \citep{Makarevich2016a}.
\begin{equation}
	\label{eqn:gdi_Mak16}
	(H_i-H_e)\omega = (H_i\vec{V}_{e0}-H_e\vec{V}_{i0})\cdot\vec{k}+(C_i-C_e)H_eH_i
\end{equation}
As before, \(\vec{k}\) is the wavevector and \(\vec{V}_{\alpha 0}\) is the velocity of species \(\alpha\).  The species \(\alpha\) is given by either \(i\) for ions or \(e\) for electrons.  The other quantities in equation \ref{eqn:gdi_Mak16} are not standard and described below.
\begin{equation}
	H_\alpha = S_\alpha F_\alpha + D_\alpha^{-1} F_\parallel \quad\quad\quad
	C_\alpha = \frac{T_\alpha}{m_\alpha \Omega_\alpha}
\end{equation}
\begin{equation}
	F_\alpha = i k_\perp^2 D_\alpha + \vec{G}\cdot\vec{k}_\perp D_\alpha + \vec{G}\cdot\vec{k}\times\uvec{b} \quad\quad\quad
	F_\parallel = i k_\parallel^2 + \vec{G}\cdot\vec{k}_\parallel
\end{equation}
\begin{equation}
	S_\alpha = \frac{1}{1+D_\alpha^2} \quad\quad\quad
	D_\alpha = -\frac{i}{\Omega_\alpha}(\omega-\vec{k}\cdot{V}_{\alpha 0})+r_\alpha
\end{equation}
The quantity \(r_\alpha\) is the ratio of the collision frequency to the gyrofrequency of a particular species, given by \(r_\alpha = \nu_\alpha/\Omega_\alpha\).

@@

By considering certain limiting cases, an expression that is more relevant to the polar F region specifically can be resolved from Equation \ref{eqn:gdi_Mak16_alt}.  First, assume a cold plasma such that the temperature of both ions and electrons is negligible.  In this case, \(T_\alpha = 0\), so \(C_\alpha = 0\).  Secondly assume all irregularities are perfectly aligned with the magnetic field such that \(k_\parallel = 0\).  This means that \(\vec{k}_\perp = \vec{k}\), \(y=0\), and \(Y=0\).  Finally, consider only low frequencies (\(\omega \ll \Omega_\alpha\)) and long wavelengths (\(\vec{k}\cdot\vec{V}_{\alpha 0} \\ \nu_\alpha\)).  If \(D_\alpha\) is expressed as shown below, it can bee seen that these limits result in the first two terms being negligible such that \(D_\alpha \approx r_\alpha\).
\begin{equation}
	D_\alpha = -i\frac{\omega}{\Omega_\alpha} + i\frac{\vec{k}\cdot\vec{V}_{\alpha 0}}{\Omega_\alpha} + \frac{\nu_\alpha}{\Omega_\alpha} \approx \frac{\nu_\alpha}{\Omega_\alpha} = r_\alpha
\end{equation}
Equation \ref{eqn:gdi_Mak16_alt} can then be expressed as follows.
\begin{equation}
	\label{eqn:gdi_Mak16_Freg}
	(r_i-r_e)\omega = \vec{V}_d\cdot\vec{k}\left(\frac{Z_2}{Z_1}\right)+(r_i\vec{V}_{i0}-r_e\vec{V}_{e0})
\end{equation}
The growth rate, \(\gamma\), is contained within the dispersion relation as \(\omega = \omega_r+i\gamma\), so the imaginary part of Equation \ref{eqn:gdi_Mak16_Freg} must be considered separately.
\begin{equation}
	\label{eqn:gdi_Mak16_gam}
	(r_i-r_e)\gamma = \vec{V}_d\cdot\vec{k}\;\text{Im}\left\{\frac{Z_2}{Z_1}\right\}
\end{equation}
The differential plasma drift has been defined previously in Equation \ref{eqn:gdi_Mak16_Vd} but the imaginary component of \(Z_2/Z_1\) must still be identified.  Using the definitions of \(Z_1\) and \(Z_2\) provided in Equation \ref{eqn:gdi_Mak16_Z} and the simplifying assumptions discussed above for the F region,
\begin{equation}
	\frac{Z_2}{Z_1} = \frac{R(i+a)-b}{i+a+Rb}
\end{equation}.
Multiplying both the numerator and denominator by \(-i+a+Rb\) removes the imaginary component from the denominator.  Additionally the local approximation \(G/k_\perp \ll 1\) is taken, which results in \(a\), \(b\), and \(c\) being small so that any second order terms in these quantities can be neglected.
\begin{equation}
	\frac{Z_2}{Z_1} = R+ib(1+R^2)
\end{equation}
Using the above result and Equation \ref{eqn:gdi_Mak16_Vd}, Equation \ref{eqn:gdi_Mak16_gam} becomes
\begin{equation}
	(r_i-r_e)\gamma = (r_i-r_e)s_es_i\left(1+\psi\right)\left(R\vec{V}_E-\frac{E}{B}\right)\cdot\vec{k}b\left(1+R^2\right)
\end{equation}
Recognizing that \(s_es_i\left(1+R^2\right) = (1+\psi)^{-2}\) and \(b = \vec{G}\cdot\vec{k}\times\uvec{b}/k^2 = -\vec{k}\cdot\uvec{b}\times\vec{G}/k^2\) by vector identities a final expression for the GDI growth rate in the F region can be found, which agrees well with the results of \citet{Makarevitch2014c}.
\begin{equation}
	\label{eqn:gdi}
	\gamma = \frac{1}{1+\psi}\left(\uvec{k}\cdot\uvec{b}\times\vec{G}\right)\uvec{k}\cdot\left(\frac{\vec{E}}{B}-R\vec{V}_E\right)
\end{equation}

@@

\subsection{Observations}

Polar patches were first observed in the 1980s using optical and radio techniques \citep{Weber1981,Buchau1983,Weber1984}.  \citet{Weber1981} identified sun-aligned arcs that extend  along the sun-earth line through large parts of the polar cap using all-sky imaging photometers (ASIP).  These arcs are over 1000 km long and \(\sim100\) km wide and are produced by soft electron precipitation.  \citet{Buchau1983} used a combinations of Digisonde and ASIP measurements to identify large luminous patches in the polar cap in addition to sun-aligned arcs.  These patches could have densities up to \(10^6\) cm\(^{-3}\) and tended to drift antisunward.  \citet{Weber1984} found patches of enhanced ionization between 800--1000 km during moderately disturbed geomagnetic conditions, again using a combination of ASIP images, ionosonde measurements, and data from Dynamic Explorer 2 (DE-2).  These finings confirmed that density enhancements tended to drift antisunward at speeds of 500-1000 m/s and patches were not colocated with enhanced particle precipitation within the polar cap, indicating that this is not the source of the density enhancement.  Over time, the trailing edge of a patch became steeper than the leading edge and irregularities developed on the trailing edge.

\citet{Coley1998} examined the occurrence of polar patches in both the northern and southern hemispheres with DE-2 and found that the frequency of patch occurrence was greatest when  the cusp was slightly dayward of the terminator.  This creates the situation of an unsunlit polar cap, so background plasma density is low and blobs of enhanced density are easily visible.  In addition, the cusp then provides a gateway of highly ionized sunlit plasma into the dark polar cap.  \citet{Rodger1996} also considered occurrence of patches but with a HF radar in Halley, Antarctica.  The diurnal variation in patch occurrence was found to peak at magnetic noon and the seasonal variation in equinox months.  The only correlation between hourly averaged solar wind parameters that was found was a bias towards patches occurring with a negative IMF \(B_z\) component.

The theoretical mechanism for patch formation via transient magnetic reconnection has been described previously, Section \ref{sec:patch_formation}.  \citet{Carlson2004} has provided experimental evidence to support this mechanism.  Five signatures of transient magnetic reconnection were identified and then a patch formation event was examined for these signatures using the European Incoherent Scatter Scientific Association (EISCAT) Svalbard Radar.  All five signatures were observed as expected, strengthening the idea that transient magnetopause reconnection is responsible for patch formation.  \citet{Carlson2006} furthered this using a observations from the EISCAT Svalbard Radar, the EISCAT VHF radar in Troms\o, Norway, an ASIP, and a meridian scanning photometer (MSP).  They managed to identify and track a series of patches directly from the subauroral plasma reservoir and show that the boundary moves equatorward before relaxing poleward, one of the main predictions of the transient magnetopause reconnection mechanism \citep{Lockwood1992b}.

@@

It is known that the polar cap ionosphere is highly structured in a non-trivial manner \citep{Carlson2012}.  In addition to being of purely scientific interest, this structuring creates complications when sending radio waves through the ionosphere.  Because the structuring is highly irregular and varies on many spatial and temporal scales, unpredictable wave refraction, phase shifts, and amplitude attenuation can occur, making signals difficult to detect.  This introduces problems in any system that involves ground-to-satellite communication, such as navigation or communication.  Because of the increasing interconnectedness of modern society, any interruptions or errors in these systems can have broad impacts on infrastructure, including commercial and defense interests, as well as everyday life.

>>>>

\section{A Brief Review of Plasma Structuring Theory and Observations}
Much work has been done over the past several decades to study the complicated picture of plasma structuring in the polar cap, including theoretical analysis of plasma dispersion relations, simulations of plasma dynamics, and observations of plasma structuring using a variety of instruments.  Here, a survey will be given of some of these advancements and the current state of understanding plasma structring on a range of scales.

\subsection{Theory and Simulations}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here we will focus on large regions of enhanced plasma density known as polar patches  \citep[][e.g.]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  The first observations of polar patches were made in the 1960s by \citet{Hill1963}.  They can range in size up to 1500 km in diameter and can have peak densities as much as eight times that of the background ionospheric plasma \citep{Weber1986,Hosokawa2016}.  Polar patches are know to emerge on the dayside and then propagate with the background plasma convection across the polar cap to the nightside, where they either disintegrate or recombine with enhanced-density plasma in the auroral oval \citep{Weber1985,Weber1986}.

There have been at least 6 mechanisms proposed for how polar patches form on the dayside \citep{Crowley1996,Carlson2012}.
\begin{enumerate}
	\item Discrete changes in solar wind parameters, usch as IMF, density, speed, and pressure \citep{Sojka1994}
	\item Average flow patterns varying in time \citep{Anderson1988}
	\item Transient magnetopause reconnection \citep{Lockwook1992b}
	\item Plasma production by cusp particle precipitation \citep{Rodger1994,Millward1999}
	\item Plasma flow jet channels that cut continuous streaches of plasma into segments 	\citep{Valladares1998}
	\item Alfven wave coupling \citep{Prikryl1999}
\end{enumerate}
Although particle precipitation has been shown to create enhanced plasma density in the cusp region \citep{Roger1994}, it is unreasonable for this mechanism to account for the largest density enhancements observed in polar patches.  These high densities must originate from the resivor of solar illuminated plasma on the dayside.  Additionally, a mechanism is still required to separate patches from the cusp region.  High speed flow channels can ``cut'' segments off of a tongue of high-density plasma that has been pulled into the polar cap by convection \citep{Valladares1994,Valladares1998}.  Models have also shown changing large scale convection patterns quickly in time does produce density structures similar to patches \citep{Anderson1988}.  However, the mechanism that seems to be dominant for the majority of polar patches observed is transient magnetic reconnection \citep{Carlson2012}.

In order to discuss patch formation by transient magnetic reconnection, it is important to first understand the steady-state ionospheric convection patterns in the polar cap and how they relate to the magnetosphere \citep{Cowley1980}.  Figure \ref{fig:magnetosphere} is a schematic of how the IMF, magnetosphere, and ionospheric polar cap are all interconnected.  In general, magnetic field lines in the magnetosphere can either be closed (grey lines in Figure \ref{fig:magnetosphere}), meaning they connect only the north and south pole of the earth, or open (light blue lines in Figure \ref{fig:magnetosphere}), when one of the ends of the field lines originates from the sun such that the field line is connected directly to the IMF.  IMF lines that are not connected to the earth are shown in dark blue in figure \ref{fig:magnetosphere}.  The polar cap boundary is generally defined as the border between open and closed magnetic field lines, shown by the green line in Figure \ref{fig:magnetosphere}.

By Faraday's Law, the total electromotive force, \(\xi\) around the polar cap boundary is equal to the rate of change of magnetic flux, \(\Phi_B\) within the polar cap, Equation \ref{eqn:pc_imf} \citep{Lockwood1992a}.  In steady-state solutions, \(d\Phi_B/dt = 0\), so \(\xi=0\) by definition, however the polar cap is known to be a highly dynamic system, so changes in magnetic flux are expected.
\begin{equation}
	\label{eqn:pc_imf}
	-\xi = \frac{d\Phi_B}{dt} = B\frac{dA_{pc}}{dt}
\end{equation}
In the polar cap at ionospheric altitudes, the magnetic field strength is relatively constant at around \(5\times 10^-5\) T, so any changes in magnetic flux must be due to changes in the polar cap area, \(A_{pc}\) due to the polar cap boundary expanding or contracting \citep{Lockwood1992a}.  The polar cap boundary changes in response to magnetic reconnection between the magnetosphere and the IMF  on both the day and night sides changing the amount of open magnetic flux.

@@

Magnetic reconnection between the IMF and the Earth's magnetosphere occurs along the X-line on the dayside magnetopause (red line in figure \ref{fig:magnetosphere}).  The X-line in the magnetopuase can be mapped along the magnetic field lines (orange) to the ionosphere, where it is referred to as the merging gap (pink dotted line in figure \ref{fig:magnetosphere}) and lies along the steady state polar cap boundary.

During reconnection events, the polar cap boundary moves equatorward as the open magnetic flux into the polar cap increases, Figure \ref{fig:magnetosphere} \citep{Cowley1991,Lockwood1992a}.  In addition, the potential along the X-line also gets mapped to the merging gap, assuming there is minimal drop in potential along the magnetic field lines \citep{Lockwood1992b}.  The combination of these two effects results in flux being transferred across the boundary at a rate equal to the applied voltage  along the X-line and hence plasma flow across the merging gap \citep{Lockwood1992b}, as seen in Figure \ref{fig:patch_formation}.

Figure \ref{fig:patch_formation} is a schematic of how transient magnetopause reconnection forms polar patches.  In all panels, the the dayside is the top of the figure and the green line shows the open-closed polar cap boundary, similar to Figure \ref{fig:magnetosphere}.  Plasma flow contours are shown in black and the light orange region represents high density from the dayside reservoir.  The reconnection burst starts at Figure \ref{fig:patch_formation}a and the increasing amount of open magnetic flux moves the dayside open-closed boundary equatorwards, as described above \citep{Cowley1991}.

As reconnection continues in Figure \ref{fig:patch_formation}b, the open-closed boundary continues to expand equatorwards as more open flux is produced, which excites a convection pattern in the polar cap \citep{Cowley1991}.  Simultaneously, this convection starts to transport dense dayside plasma polewards \citep{Lockwood1992b}.  As more open flux is produced, Figure \ref{fig:patch_formation}c, the strength of the plasma flow increases and a large blob of dense, dayside plasma is pulled into the polar cap \citep{Lockwood1992b}.

At Figure \ref{fig:patch_formation}d, the burst of reconnection stops and the open-closed boundary begins to relax polewards \citep{Cowley1991}.  As it moves, it pulls the blob of dense plasma within the polar cap with it, separating it from the reservoir on the dayside \citep{Lockwood1992b}.  Before the convection flow stops completely, Figure \ref{fig:patch_formation}e, the blob of dense plasma ``pinches off'' from the dayside plasma to become an independant patch \citep{Lockwood1992b}.  The exact mechanism by which this happens is not well understood, but it is thought to be related to small variations in the IMF By component shifting the convection pattern slightly \citep{Cowley1980,Lockwood1992b}.  By Figure \ref{fig:patch_formation}f, the open-closed boundary has returned completely to its original position and there is a newly formed patch within in the polar cap.  The process can now repeat to form another patch within the polar cap.

A series of reconnection bursts can create a line of patches all propagating across the polar cap.  Reconnection bursts typically last about 2 minutes with anywhere between 7 and 25 minutes beween bursts \citep{Foster1984,Etemadi1988,Lockwood1992b}.  However, observations have shown convection flow to be close to continuous for much longer periods.  This can be explained by each reconnection burst exciting flow for a much longer interval than the time between bursts.  In this way, a reconnection burst can excite plasma flow before the convection from the previous burst has stopped, and a series of overlapping reconnection events like this can create continuous plasma flow through the polar cap \citep{Cowley1991}.

@@

Plasma structuring, especially at smaller scales, is often discussed in terms of the growth or damping of plasma waves, or instabilities.  As discussed previously in Section \ref{sec:ionosphere_regions}, the plasma characteristics change with altitude in the ionosphere, which causes a variety of different instability mechanisms to be operational depending on the region considered.  The Farley-Buneman instability (FBI), or the modified two-stream instability tends to be a major factor in the E region where ion inertial effects are high \citep{Farley1963,Buneman1963}.  Also operational in the E region but more dominant at higher altitudes is the gradient-drift instability (GDI) \citep{Simon1963,Hoh1963,Linson1970}.  If a wave has a component along the magnetic filed line, the current-conductive instability (CCI) can be operational even if GDI is stable.  Addition instabilities mechanisms emerge if shears are considered, when \(\nabla E \neq 0\), such as the Kelvin-Helmholtz instability (KHI).

@@

Some of the first 1D investigations of the linear GDI growth rate found that it can be described by a simple expression involving the plasma drift velocity, \(V_E = \vec{E}\times\vec{B}\), and the gradient scale length, \(L = |n/(\partial n/\partial x)|\), Equation \ref{eqn:gdi_old} \citep{Simon1963,Hoh1963,Linson1970}.
\begin{equation}
	\label{eqn:gdi_old}
	\gamma = \frac{V_E}{L}
\end{equation}
This expression is only valid if the density gradient is in the same direction as the plasma drift velocity and the wave is propagating perpendicular to both.  The results of \citet{Keskinen1982,Keskinen1983}, Equation \ref{eqn:gdi_80s}, do consider an arbitrary wavevector, \(\vec{k}\), explicitly to some extent, but still assume a general relationship between density gradients and plasma drift \citep{Tsunoda1988}.
\begin{equation}
	\label{eqn:gdi_80s}
	\gamma = \frac{k_y}{k}\frac{\vec{k}\cdot\vec{V}_E}{kL}
\end{equation}
The directional dependance of GDI has been shown to be significant, such that the growth rate can actually change signs (determining whether wave growth or damping occurs) with different wavevector directions, so it is important to consider this factor \citep{Makarevich2014c}.  In addition, these expressions have historically only applicable in a particular altitudinal regieme, usually the F region.

Recently, a new expression has been introduced that considers any arbitrary wavevector, density gradient, and altitude within the ionosphere, which allows FBI, GDI, an CCI to be considered with a single dispersion relation \citep{Makarevich2016a}.
\begin{equation}
	\label{eqn:gdi_Mak16}
	(H_i-H_e)\omega = (H_i\vec{V}_{e0}-H_e\vec{V}_{i0})\cdot\vec{k}+(C_i-C_e)H_eH_i
\end{equation}
As before, \(\vec{k}\) is the wavevector, \(\omega\) is the wave frequency, and \(\vec{V}_{\alpha 0}\) is the velocity of species \(\alpha\).  The species \(\alpha\) is given by either \(i\) for ions or \(e\) for electrons.  The other quantities in equation \ref{eqn:gdi_Mak16} are not standard and described below.
\begin{equation}
	H_\alpha = S_\alpha F_\alpha + D_\alpha^{-1} F_\parallel \quad\quad\quad 
	C_\alpha = \frac{T_\alpha}{m_\alpha \Omega_\alpha}
\end{equation}
\begin{equation}
	F_\alpha = i k_\perp^2 D_\alpha + \vec{G}\cdot\vec{k}_\perp D_\alpha + \vec{G}\cdot\vec{k}\times\uvec{b} \quad\quad\quad
	F_\parallel = i k_\parallel^2 + \vec{G}\cdot\vec{k}_\parallel
\end{equation}
\begin{equation}
	S_\alpha = \frac{1}{1+D_\alpha^2} \quad\quad\quad
	D_\alpha = -\frac{i}{\Omega_\alpha}(\omega-\vec{k}\cdot{V}_{\alpha 0})+r_\alpha
\end{equation}
The quantity \(r_\alpha\) is the ratio of the collision frequency to the gyrofrequency of a particular species, given by \(r_\alpha = \nu_\alpha/\Omega_\alpha\).

@@

By considering certain limiting cases, an expression that is more relevant to the polar F region specifically can be resolved from Equation \ref{eqn:gdi_Mak16_alt}.  First, assume a cold plasma such that the temperature of both ions and electrons is negligible.  In this case, \(T_\alpha = 0\), so \(C_\alpha = 0\).  Secondly assume all irregularities are perfectly aligned with the magnetic field such that \(k_\parallel = 0\).  This means that \(\vec{k}_\perp = \vec{k}\), \(y=0\), and \(Y=0\).  Finally, consider only low frequencies (\(\omega \ll \Omega_\alpha\)) and long wavelengths (\(\vec{k}\cdot\vec{V}_{\alpha 0} \gg \nu_\alpha\)).  If \(D_\alpha\) is expressed as shown below, it can bee seen that these limits result in the first two terms being negligible such that \(D_\alpha \approx r_\alpha\).
\begin{equation}
	D_\alpha = -i\frac{\omega}{\Omega_\alpha} + i\frac{\vec{k}\cdot\vec{V}_{\alpha 0}}{\Omega_\alpha} + \frac{\nu_\alpha}{\Omega_\alpha} \approx \frac{\nu_\alpha}{\Omega_\alpha} = r_\alpha
\end{equation}
Equation \ref{eqn:gdi_Mak16_alt} can then be expressed as follows.
\begin{equation}
	\label{eqn:gdi_Mak16_Freg}
	(r_i-r_e)\omega = \vec{V}_d\cdot\vec{k}\left(\frac{Z_2}{Z_1}\right)+(r_i\vec{V}_{i0}-r_e\vec{V}_{e0})
\end{equation}
The growth rate, \(\gamma\), is contained within the dispersion relation as \(\omega = \omega_r+i\gamma\), so the imaginary part of Equation \ref{eqn:gdi_Mak16_Freg} must be considered separately.
\begin{equation}
	\label{eqn:gdi_Mak16_gam}
	(r_i-r_e)\gamma = \vec{V}_d\cdot\vec{k}\;\text{Im}\left\{\frac{Z_2}{Z_1}\right\}
\end{equation}
The differential plasma drift has been defined previously in Equation \ref{eqn:gdi_Mak16_Vd} but the imaginary component of \(Z_2/Z_1\) must still be identified.  Using the definitions of \(Z_1\) and \(Z_2\) provided in Equation \ref{eqn:gdi_Mak16_Z} and the simplifying assumptions discussed above for the F region,
\begin{equation}
	\frac{Z_2}{Z_1} = \frac{R(i+a)-b}{i+a+Rb}
\end{equation}.
Multiplying both the numerator and denominator by \(-i+a+Rb\) removes the imaginary component from the denominator.  Additionally the local approximation \(G \ll k_\perp\) is taken, which results in \(a\), \(b\), and \(c\) being small so that any second order terms in these quantities can be neglected.
\begin{equation}
	\frac{Z_2}{Z_1} = R+ib(1+R^2)
\end{equation}
Using the above result and Equation \ref{eqn:gdi_Mak16_Vd}, Equation \ref{eqn:gdi_Mak16_gam} becomes
\begin{equation}
	(r_i-r_e)\gamma = (r_i-r_e)s_es_i\left(1+\psi\right)\left(R\vec{V}_E-\frac{E}{B}\right)\cdot\vec{k}b\left(1+R^2\right)
\end{equation}
Recognizing that \(s_es_i\left(1+R^2\right) = (1+\psi)^{-2}\) and \(b = \vec{G}\cdot\vec{k}\times\uvec{b}/k^2 = -\vec{k}\cdot\uvec{b}\times\vec{G}/k^2\) by vector identities, a final expression for the GDI growth rate in the F region can be found, which agrees well with the results of \citet{Makarevitch2014c}.
\begin{equation}
	\label{eqn:gdi}
	\gamma = \frac{1}{1+\psi}\left(\uvec{k}\cdot\uvec{b}\times\vec{G}\right)\uvec{k}\cdot\left(\frac{\vec{E}}{B}-R\vec{V}_E\right)
\end{equation}

@@

\subsection{Observations of Polar Cap Structuring}

Polar patches were first observed in the 1960s \citep{Hill1963}, but much of the work of characterizing them using optical and radio techniques was not accomplished until two decades later \citep{Weber1981,Weber1984,Weber1986,Buchau1983,Buchau1985}.  \citet{Weber1981} identified sun-aligned arcs that extend  along the sun-earth line through large parts of the polar cap using all-sky imaging photometers (ASIP).  These arcs are over 1000 km long and \(\sim100\) km wide and are produced by soft electron precipitation.  \citet{Buchau1983} used a combinations of Digisonde and ASIP measurements to identify large luminous patches in the polar cap in addition to sun-aligned arcs.  These patches could have densities up to \(10^6\) cm\(^{-3}\) and tended to drift antisunward.  Additionally, \citet{Buchau1983} found that polar patches were more likely to be observed for southward IMF \(B_z\) while sun-aligned arcs were more common for northwards \(B_z\).  \citet{Weber1984} found patches of enhanced ionization between 800--1000 km during moderately disturbed geomagnetic conditions, again using a combination of ASIP images, ionosonde measurements, and data from Dynamic Explorer 2 (DE-2).  These findings confirmed that density enhancements tended to drift antisunward at speeds of 500-1000 m/s and patches were not colocated with enhanced particle precipitation within the polar cap, indicating that this is not the source of the density enhancement.  Over time, the trailing edge of a patch became steeper than the leading edge and irregularities developed on the trailing edge.  Conversely, \citet{Buchau1985} found that plasma densities within patches were comparable with those measured at dayside sub-auroral latitudes, suggesting that dayside solar-illuminated plasma is the source of dense plasma that form polar patches.

\citet{Coley1998} examined the occurrence of polar patches in both the northern and southern hemispheres with DE-2 and found that the frequency of patch occurrence was greatest when  the cusp was slightly dayward of the terminator.  This creates the situation of an unsunlit polar cap, so background plasma density is low and blobs of enhanced density are easily visible.  In addition, the cusp then provides a gateway of highly ionized sunlit plasma into the dark polar cap.  Several studies had previously reported density enhanced patches convecting from the dayside through the cusp/throat region toward the polar cap \citep{Kelley1984,Foster1984,Foster1985,Foster1993}.  This also agrees with observations made by \citet{Sojka1982,delaBeaujardiere1985}  \citet{Rodger1996} also considered occurrence of patches but with a HF radar in Halley, Antarctica.  The diurnal variation in patch occurrence was found to peak at magnetic noon and the seasonal variation in equinox months.  The only correlation between hourly averaged solar wind parameters that was found was a bias towards patches occurring with a negative IMF \(B_z\) component.  A review of historical observations of polar patches has been provided by \citet{Crowley1996}

The theoretical mechanism for patch formation via transient magnetic reconnection has been described previously, Section \ref{sec:patch_formation}.  A variety of experimental evidence has been presented to support this mechanism \citep{Cowley1998,Carlson2002}.  Experiments run in the 1980s using the European Incoherent Scatter Scientific Association (EISCAT) Troms\o, Norway ISR showed that plasma flow occurred near noon shortly after southward IMF \(B_z\) arrives at the magnetopause \citep{Etemadi1988,Todd1988}.  Later, \citet{Lockwood1993a,Lockwood1993b} showed that these plasma flows are pulsed in the region of the polar cap boundary, as predicted by \citet{Cowley1991}.  Observations made by an ASIP in Svalbard in 1984 are fully consistent with patch formation through transient magnetopause reconnection \citep{Carlson1996,Carlson2002}.  \citet{Carlson2004} identified five signatures of transient magnetic reconnection and then examined a patch formation event for these signatures using the EISCAT Svalbard Radar.  All five signatures were observed as expected, strengthening the idea that transient magnetopause reconnection is responsible for patch formation.  \citet{Carlson2006} furthered this using a observations from the EISCAT Svalbard Radar, the EISCAT VHF radar in Troms\o an ASIP, and a meridian scanning photometer (MSP).  They managed to identify and track a series of patches directly from the subauroral plasma reservoir and show that the boundary moves equatorward before relaxing poleward, one of the main predictions of the transient magnetopause reconnection mechanism \citep{Lockwood1992b}.

@@

It is well accepted that the polar cap ionosphere is highly structured in a non-trivial manner \citep{Carlson2012}.  In addition to being of purely scientific interest, this structuring creates complications when sending radio waves through the ionosphere.  Because the structuring is highly irregular and varies on many spatial and temporal scales, unpredictable wave refraction, phase shifts, and amplitude attenuation can occur, making signals difficult to detect.  This introduces problems in any system that involves ground-to-satellite communication, such as navigation or communication.  Because of the increasing interconnectedness of modern society, any interruptions or errors in these systems can have broad impacts on infrastructure, including commercial and defense interests, as well as everyday life.
